A Simple Type Theory for Category Theory

### 2022-04-25 paper/paper.tex
A straightforward presentation of the models. Eureka!

First, we recount the definition of a virtual double category.
\begin{definition}[Virtual Double Category]
  A (vertically strict) virtual equipment $\mathcal V$ consists of
  \begin{enumerate}
    \item A set $\mathcal V_0$ of objects.
    \item For each $\cat C, \cat D \in V_0$, a set $V_v(\cat C, \cat
      D)$ of vertical arrows, with composition and unit operations
      satisfying the category axioms.
    \item A set $V_h$ of horizontal arrows, with $d^+,d^- : V_h \to
      V_0$.

    \item For any vertical arrows $f \in V_v(\cat C, \cat D)$ and $g
      \in V_v(\cat E, \cat F)$ and horizontal arrow $R \in V_h$ with
      $d^+R = \cat F$ and $d^-R = \cat D$, a \emph{restriction}
      $R(f,g)$ with $d^+R = \cat C$ and $d^-R = \cat E$.

      This operation is furthermore associative and unital $R(f \circ
      f',g \circ g') = R(f,g)(f',g')$ and $R(\id,\id) = R$.

    \item For each composable string $\Phi$ (defined below) of horizontal arrows
      starting at $\cat C$ and ending at $\cat D$, and horizontal
      arrow $R$ with $d^-R = \cat C$ and $d^+R = \cat D$, a set
      $V_2(\Phi; R)$ of 2-cells.

    \item For any string of vertical arrows $\vec f$ into $\Phi$ and
      2-cell $r \in \mathcal V_2(\Phi,S)$, a restriction operation
      $r(\vec f) \in \mathcal V_2(\Phi(\vec f), S(d^-(\vec f),
      d^+(\vec f)))$. Furthermore, this is associative and unital
      $r(\vec id) = r$ and $r(\vec f \circ \vec g) = r(\vec {f \circ
        g})$

    \item For each $R \in V_h$, an identity arrow $\id_R \in
      V_2(d^-R,R,d^+ R; R)$ with $d^\pm \id_R = \id_{d^\pm R}$.

    \item For any $2$-cell $\alpha \in V_2(\Phi;R)$ and composable
      string $\phi$ of 2-cells into $\Phi$, a composite $\alpha \circ
      \phi$.

    \item Composition should be unital in that:
      \[ \id_R \circ [d^-\alpha, \alpha, d^+\alpha] = \alpha  \]
      and
      \[ \alpha \circ [\vec \id] = \alpha\]
    \item And associative in that
      \[ (\alpha \circ [\beta_1,\ldots]) \circ [\gamma_{11},\ldots,\gamma_{21},\ldots,] = \alpha \circ [(\beta_1 \circ \gamma_{11},\ldots), (\beta_2 \circ \gamma_{21},\ldots)]\]
  \end{enumerate}

  A composable string $\Phi$ of horizontal arrows is defined to be an
  alternating sequence of objects and horizontal arrows, beginning and
  ending with an object such that any substring $\cat C,R,\cat D$
  satisfies $d^-R = \cat C$ and $d^+R = \cat D$.

  A composable string of vertical arrows $\vec f$ into $\Phi$ is a
  sequence of vertical arrows, one for each object in $\Phi$, with the
  codomain of the arrow matching the object in $\Phi$. Then

  A composable string of 2-cells from $\Phi$ to $\Psi$ is a sequence
  $\vec r$ of 2-cells, one for each horizontal arrow in $\Psi$ such
  that the codomains match the objects in
\end{definition}

Our main departure from\cite{crutwellshulman} is making restrictions
part of the definition, and also requiring them to be strictly
associative and unital. In the examples that we are most concerned
with, however, these operations are strict and so we will not explore
the possibility of coherence theorems for proving this equivalent to a
weaker definition.

\begin{definition}
  A functor of virtual double categories $F : \mathcal V \to \mathcal
  U$ is a mapping on objects, vertical arrows, horizontal arrows and
  2-cells respecting typing and preserving composition, units, and
  restriction.
\end{definition}

>>>>

%% 1. A strict double category is a monoid in Span(Cat)
%% 2. A strict double category internal to a topos T is a monoid in Span(T-Cat)
%% 3. A double category with display pro-arrows is a strict double category with a span of profunctors that is representable
%%    C0 <- C1 -> C0
%% 4. UMPs are given on top of that structure

We build up the notion of a model in several stages.

\begin{definition}
  A strict double category is a monoid in the double category $Span(Cat)$.
\end{definition}

This data consists of a span $C_0 \leftarrow C_1 \rightarrow C_0$
where $C_0$ is the vertical category and $C_1$ is the category whose
objects are horizontal morphisms and morphisms are squares, with the
two projections giving the left and right ``frame'' for the horizontal
morphisms/squares. The unit gives the horizontal identity morphisms
and the monoid multiplication is the horizontal composition.

\begin{definition}
  A double category with displays (DCD) consists of
  \begin{enumerate}
  \item A strict double category $C$
  \item A category $C_d$ of (horizontal) ``display morphisms'' and ``display squares''.
  \item A span of profunctors:
    % https://q.uiver.app/?q=WzAsOCxbMiwyLCJDXzEiXSxbMCwyLCJDXzAiXSxbNCwyLCJDXzAiXSxbMiwwLCJDX2QiXSxbMCwwLCJDXzAiXSxbNCwwLCJDXzAiXSxbMSwxLCJcXExlZnRhcnJvdyJdLFszLDEsIlxcUmlnaHRhcnJvdyJdLFszLDAsIkNfXFx2ZGFzaCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9fX1dLFs1LDIsIlxcdGV4dHJte0hvbX0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifX19XSxbNCwxLCJcXHRleHRybXtIb219IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV0sWzMsNCwicyIsMl0sWzMsNSwidCJdLFswLDIsInQiLDJdLFswLDEsInMiXV0=
\[\begin{tikzcd}
	{C_0} && {C_d} && {C_0} \\
	& \Leftarrow && \Rightarrow \\
	{C_0} && {C_1} && {C_0}
	\arrow["{C_\vdash}"', "\shortmid"{marking}, from=1-3, to=3-3]
	\arrow["{\textrm{Hom}}", "\shortmid"{marking}, from=1-5, to=3-5]
	\arrow["{\textrm{Hom}}"', "\shortmid"{marking}, from=1-1, to=3-1]
	\arrow["s"', from=1-3, to=1-1]
	\arrow["t", from=1-3, to=1-5]
	\arrow["t"', from=3-3, to=3-5]
	\arrow["s", from=3-3, to=3-1]
\end{tikzcd}\]
\item A fully faithful functor $i : C_d \to C_1$ that respects source and target in that the following commute:
% https://q.uiver.app/?q=WzAsNixbMiwyLCJDXzEiXSxbMCwyLCJDXzAiXSxbNCwyLCJDXzAiXSxbMiwwLCJDX2QiXSxbMCwwLCJDXzAiXSxbNCwwLCJDXzAiXSxbMywwLCJpIiwyXSxbNSwyLCJcXGlkIl0sWzQsMSwiXFxpZCIsMl0sWzMsNCwicyIsMl0sWzMsNSwidCJdLFswLDIsInQiLDJdLFswLDEsInMiXV0=
\[\begin{tikzcd}
	{C_0} && {C_d} && {C_0} \\
	\\
	{C_0} && {C_1} && {C_0}
	\arrow["i"', from=1-3, to=3-3]
	\arrow["\id", from=1-5, to=3-5]
	\arrow["\id"', from=1-1, to=3-1]
	\arrow["s"', from=1-3, to=1-1]
	\arrow["t", from=1-3, to=1-5]
	\arrow["t"', from=3-3, to=3-5]
	\arrow["s", from=3-3, to=3-1]
\end{tikzcd}\]
and $i$ represents $C_\vdash$ in that there is a natural isomorphism:
    \[ e : C_\vdash(\Phi, R) \cong C_1(\Phi, i(R)) \]
    that commutes with the natural transformations $s,t$.
  \end{enumerate}
\end{definition}

In terms of the syntax of SCT, the category $C_0$ models the category
and functor judgments, the category $C_1$ models the profunctor
\emph{contexts} and substitutions and the category $C_d$ models the
profunctor judgment. The profunctor $C_\vdash$ models the
transformation judgment and the functor $i$ models the construction of
a ``singleton'' context $\alpha:\cat C,x:R,\beta:\cat D$.

A DCD is a more general notion that a VDC, but they are closely
related, in that VDCs form a coreflective subcategory of DCDs. This is
analogous to the situation in models of dependent type theory, where
DCDs are analogous to Categories with Families \cite{cwfs,natmodels}
and VDCs are analogous to the more ``syntactic'' notion of a
Contextual Category/C-system \cite{ctxcat,csystem}. First, any DCD
induces a VDC by taking the display proarrows as the proarrows and
interpreting 2-cells using $i$ and $C_\vdash$.  Second, any VDC
presents a DCD whose double category has the original VDC pro-arrows
as the display morphisms and \emph{composable strings} of proarrows as
horizontal morphisms, which gives a strict double category.
%
Taking as morphisms the strict morphisms of DCDs/VDCs, these
constructions form a coreflective embedding of VDCs into DCDs.

The principle advantage of DCDs over VDCs is that the presentation of
their algebraic structure is considerably simplified and is
straightforward to describe as a Generalized Algebraic Theory in the
sense of \citet{CartmellGAT}. This is because we don't \emph{require}
that our profunctor contexts are lists, but instead merely the fact
that their concatenation forms a strict category. Then the
associativity/identity equations for a DCD are simply those of a
double category.

Next, each of the connectives of SCT are modeled by different
universal properties that are profunctor isomorphisms involving
principally $C_\vdash$.
TODO: write them down.

\begin{definition}
  An \emph{equipment with displays} (ED) is a DCD with all of the
  universal properties in Figure\ref{eq-umps}.
\end{definition}

Finally, to model the logic over SCT provided by the dependently typed
context $\Gamma$, we use a category with families $Ctx$, and interpret
all of the prior constructions in this section to be internal to
presheaves on $Ctx$. This models the syntactic aspect that all double
category structure is parameterized by a context $\Gamma$ and stable
under substitution. Since the category of presheaves is a topos, all
definitions given in set-theoretic language can be interpreted
internally.

\begin{definition}
  A Category with Families is...
\end{definition}

\begin{definition}
  A hyperdoctrine of equipment with displays (HED) is a pair of a
  category with families $Ctx$ modeling $\Sigma,\Pi,\textrm{Id}$ and a
  DCD internal to $\Set^{Ctx^o}$ with categories, functors, displays
  and unary cells representable as types.
\end{definition}

Again this notion is straightforwardly, if laboriously, interpretable
as a GAT.
